\section{Simulation Data Generating Process Setup}
\label{app:dgp_setup}
\subsection{Data Generating Process for Section~\ref{sec:simulation}}

We here provide the detailed data generating process (DGP) for the simulation study in Section~\ref{sec:simulation}. All distributions are defined on the support $[0,1]$ and we implement six different DGPs to evaluate the performance of our HAL-MLE methods across various distributional shapes and complexities.

\subsubsection{DGP Definitions and Parameters} \label{app_g_1:dgp_definitions}

\paragraph{1. Truncated Normal Distribution}
The truncated normal distribution is defined as:
\begin{align}
f(x) = \frac{\phi\left(\frac{x-\mu}{\sigma}\right)}{\sigma \left[\Phi\left(\frac{b-\mu}{\sigma}\right) - \Phi\left(\frac{a-\mu}{\sigma}\right)\right]}, \quad x \in [a,b]
\end{align}
where $\phi$ and $\Phi$ are the standard normal PDF and CDF respectively.

\textbf{Parameters:} $\mu = 0.5$, $\sigma = 0.1$, $a = 0$, $b = 1$

\paragraph{2. Sinusoidal Distribution}
The sinusoidal-based density is defined as:
\begin{align}
f(x) = \frac{\sin(\pi x) + 1.1}{C}, \quad x \in [0,1]
\end{align}
where $C = \int_0^1 (\sin(\pi t) + 1.1) dt = \frac{2}{\pi} + 1.1 \approx 1.7366$ is the normalization constant.

\textbf{Parameters:} No configurable parameters (fixed functional form)

\paragraph{3. Truncated GMM Symmetric Three Components}
A mixture of three truncated normal distributions:
\begin{align}
f(x) = \sum_{i=1}^3 w_i \cdot f_i(x), \quad x \in [0,1]
\end{align}
where each $f_i(x)$ is a truncated normal component.

\textbf{Parameters:}
\begin{itemize}
\item Component 1: $\mu_1 = 0.2$, $\sigma_1 = 0.05$, $w_1 = 0.33$
\item Component 2: $\mu_2 = 0.5$, $\sigma_2 = 0.05$, $w_2 = 0.34$
\item Component 3: $\mu_3 = 0.8$, $\sigma_3 = 0.05$, $w_3 = 0.33$
\end{itemize}

\paragraph{4. Truncated GMM Five Spikes}
A mixture of six truncated normal components with five narrow spikes and one broad component:
\begin{align}
f(x) = \sum_{i=1}^6 w_i \cdot f_i(x), \quad x \in [0,1]
\end{align}

\textbf{Parameters:}
\begin{itemize}
\item Spike components ($i = 1,\ldots,5$): $\mu_i \in \{0.45, 0.475, 0.5, 0.525, 0.55\}$, $\sigma_i = 0.005$, $w_i = 1/15$
\item Broad component: $\mu_6 = 0.5$, $\sigma_6 = 0.05$, $w_6 = 2/3$
\end{itemize}

\paragraph{5. Truncated GMM Asymmetric Three Components}
An asymmetric mixture of three truncated normal distributions:
\begin{align}
f(x) = \sum_{i=1}^3 w_i \cdot f_i(x), \quad x \in [0,1]
\end{align}

\textbf{Parameters:}
\begin{itemize}
\item Component 1: $\mu_1 = 0.35$, $\sigma_1 = 0.1$, $w_1 = 0.4$
\item Component 2: $\mu_2 = 0.65$, $\sigma_2 = 0.05$, $w_2 = 0.4$
\item Component 3: $\mu_3 = 0.9$, $\sigma_3 = 0.2$, $w_3 = 0.2$
\end{itemize}

\paragraph{6. Step Function Distribution}
A piecewise constant distribution:
\begin{align}
f(x) = \frac{1}{C} \begin{cases}
\ell_1, & x \in [0, b) \\
\ell_2, & x \in [b, 1]
\end{cases}
\end{align}
where $C = \ell_1 \cdot b + \ell_2 \cdot (1-b)$ is the normalization constant.

\textbf{Parameters:} $\ell_1 = 1.0$, $\ell_2 = 0.5$, $b = 0.7$

\subsubsection{Population Parameters}

Table~\ref{tab:dgp_population_params} summarizes the true population parameters for each DGP, including the mean, median, variance, second moment, and survival probability at $x = 0.5$.

\begin{table}[htbp]
\centering
\caption{Population Parameters for Data Generating Processes}
\label{tab:dgp_population_params}
\begin{threeparttable}
\begin{tabular}{lccccc}
\toprule
DGP & Mean & Median & Variance & Second Moment & $S(0.5)$ \\
\midrule
Truncated Normal & 0.5000$^*$ & 0.5000$^*$ & 0.010000 & 0.260000 & 0.5000$^*$ \\
Sinusoidal & 0.5000$^*$ & 0.5000$^*$ & 0.070145 & 0.320145 & 0.5000$^*$ \\
GMM Symmetric & 0.5000$^*$ & 0.5000$^*$ & 0.061896 & 0.311896 & 0.5000$^*$ \\
GMM Five Spikes & 0.5000$^*$ & 0.5000$^*$ & 0.002092 & 0.252092 & 0.5000$^*$ \\
GMM Asymmetric & 0.559669 & 0.608370 & 0.041087 & 0.354317 & 0.619610 \\
Step Function & 0.438235 & 0.425000 & 0.071283 & 0.263333 & 0.411760 \\
\bottomrule
\end{tabular}
\begin{tablenotes}
\small
\item $^*$Exact values due to symmetry around $x = 0.5$
\item $S(0.5) = P(X > 0.5)$ is the survival probability at $x = 0.5$
\end{tablenotes}
\end{threeparttable}
\end{table}

The population parameters in Table~\ref{tab:dgp_population_params} serve as ground truth for evaluating the accuracy of our HAL-MLE estimators across different distributional characteristics. 

\subsection{Random Seed Generation} \label{app_g_2:random_seed}
In this subsection, we provide the random seeds used for generating the simulation data. A master seed ($m=42$) generates a deterministic sequence of replicate seeds $\{s_r\}_{r=1}^R$ with $R=1000$ per $(\text{DGP}, n)$. In replicate $r$ the data sampler is seeded by $s_r$; cross‑validation fold assignments use the same $s_r$ (ensuring identical partitions across estimators).